### 2014-10-28 contract-language.tex
Minor edit
Adding text in transaction section
Edits to Motiv. Removed opsem. Added metatheory.

Quantifier-free propositions in our contract language are conjunctions,
disjunctions and implications of predicates expressing relations between pairs
of effect variables. The syntactic class of relations is seeded with primitive
$\visZ$, $\soZ$, and $\sameobjZ$ relations, and also admits derived relations
that are expressible as union, intersection, or transitive
closure\footnote{Strictly speaking, $R^{+}$ is not the transitive closure of
$R$, as transitive closure is not expressible in FOL. Instead, $R^{+}$ in our
language denotes \emph{a} superset of transitive closure of $R$. Formally,
$R^{+}$ is any relation $R'$ such that forall $x$, $y$, and $z$, a) $R(x,y)
\Rightarrow R'(x,y)$, and b) $R'(x,y) \conj R'(y,z) \Rightarrow R'(x,z)$} of
primitive relations. Commonly used derived relations are the \emph{same object
session order} ($\small \sooZ = \soZ ~\cap~ \sameobjZ$), the \emph{happens
before order} ($\small \hbZ = (\soZ ~\cup~ \visZ)^+$), and the per-object
happens-before order ($\small \hboZ = (\sooZ ~\cup~ \visZ)^+$).

@@

Since our contracts represent axiomatic definition of program executions, let
$\Mod{\cv}$ be the set of all executions under which contract $\cv$ is satisfied. If
every execution $\E \in \Mod{\cv_{st}}$ is also a member of $\Mod{\cv_{op}}$,
then $op$ can be safely discharged under the store consistency level
$\cv_{st}$. Formally, $\Mod{\cv_{st}} \subseteq \Mod{\cv_{op}}$. This is the
model-theoretic consequence relation, written as $\cv_{st} \models_m \cv_{op}$.

Observe that our contract language (stripped of its syntactic sugar) is a
carefully chosen subset of first-order logic that is known to be
decidable~\cite{epr}.  Completeness of the logic~\cite{completeness} allows us
to claim that $\cv_{st} \models_m \cv_{op}$ if and only if $\cv_{st}
\Rightarrow \cv_{op}$. Due to the decidability of our contract language, this
implication check is automatically discharged with the help of a theorem
prover. In this case, we say that $\cv_{op}$ is \emph{weaker than} $\cv_{st}$
(written $\cv_{op} \le \cv_{st}$).

@@

\item \textbf{Eventually consistency}: Eventually consistent operations can
  be satisfied as long as the client can reach at least one replica. For
  example, \cf{deposit} is an eventually consistent operation; its semantics
  does not require its action to manifest on all replicas before other
  operations in its session are allowed to proceed. While eventually
  consistent data store typically offer \emph{basic} eventual consistency
  with all possible anomalies, we assume that our store provides stronger
  semantics that remain highly-available~\cite{BailisHAT,COPS}; the store
  always exposes a causal cut of the updates. This semantics can be formally
  captured in terms of the following contract definition:
\vspace{-1em}
\begin{smathpar}
\ecc = \forall a,b. ~\hbo{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff}
\end{smathpar}
\noindent where $\small \hboZ = ((\soZ \cap \sameobjZ) \cup \visZ)^+$.

\item \textbf{Causal consistency}: Operations with causally consistent
  contracts are required to see a causally consistent snapshot of the object
  state, including the actions performed on the same session.  The latter
  requirement entails that if two operations $o_1$ and $o_2$ from the same
  session are applied to two different replicas $r_1$ and $r_2$, the second
  operation cannot be discharged until the effect of $o_1$ is merged with
  $o_2$ in both $r_1$ and $r_2$.  We assume that causality is only tracked
  through operations on the same object; two operations in the same session
  but on different objects are considered causally unrelated under this
  definition. Stores typically avoid tracking causality across objects to
  mitigate overheads when causality tracking is unnecessary.  The corresponding
  store semantics is captured by the contract $\ccc$ presented in
  Section~\ref{sec:store_sem}.

@@

\noindent Observe that the contracts $\scc$, $\ccc$ and $\ecc$ are
themselves totally ordered with respect to the $\le$ relation: $\ecc \le
\ccc \le \scc$. This concurs with the intuition that any contract
satisfiable under $\ecc$ or $\ccc$ is satisfiable under $\scc$ , and any
contract that is satisfiable under $\ecc$ is satisfiable under
$\ccc$. Nonetheless, we determine the class of a contract based on the
\emph{weakest} guarantee (among $\ecc$, $\ccc$, and $\scc$) required to
satisfy the contract. The classification scheme is presented formally
in Figure~\ref{sem:classify}; here, $\Delta$ captures assumptions about the nature of primitive
relations, such as $\Rvis$ and $\Rso$ are irreflexive, and the happens-before
relation $\hbZ$ is acyclic, preventing thin-air reads. \KC{Ideally, we need to
list all the axioms somewhere.}  The following definition thus precisely
captures our intuition about contract strength in terms of these rules:

\begin{definition}
A contract $\cv_{op}$ is said to be weaker than $\cv_{st}$ (written $\cv_{op}
\le \cv_{st}$ ) if and only if $\Delta \vdash \cv_{st} \Rightarrow \cv_{op}$.
\begin{center}
\end{center}
\end{definition}
\vspace{-1em}

@@

%% The rules to classify contracts into availability classes are straightforward.
%% The remaining rule judges a contract as {\sf WellFormed} if and only if it can
%% be satisfied under strong consistency. The rest are reported as errors.

>>>>

Quantifier-free propositions in our contract language are
conjunctions, disjunctions and implications of predicates expressing
relations between pairs of effect variables. The syntactic class of
relations is seeded with primitive $\visZ$, $\soZ$, and $\sameobjZ$
relations, and also admits derived relations that are expressible as
union, intersection, or transitive closure\footnote{Strictly speaking,
$R^{+}$ is not the transitive closure of $R$, as transitive closure is
not expressible in FOL.  Instead, $R^{+}$ in our language denotes
\emph{a} superset of transitive closure of $R$. Formally, $R^{+}$ is
any relation $R'$ such that forall $x$, $y$, and $z$, a) $R(x,y)
\Rightarrow R'(x,y)$, and b) $R'(x,y) \conj R'(y,z) \Rightarrow
R'(x,z)$} of primitive relations.  Commonly used derived relations are
the \emph{same object session order} ($\small \sooZ = \soZ ~\cap~
\sameobjZ$), 
%the \emph{happens before order} ($\small \hbZ = (\soZ
%~\cup~ \visZ)^+$), 
and the \emph{same object happens-before order} ($\small \hboZ = (\sooZ
~\cup~ \visZ)^+$).

@@

Since our contracts represent axiomatic definition of program
executions, let $\Mod{\cv}$ be the set of all executions under which
contract $\cv$ is satisfied. If every execution $\E \in
\Mod{\cv_{st}}$ is also a member of $\Mod{\cv_{op}}$, then $op$ can be
safely discharged under the store consistency level $\cv_{st}$.
Formally, $\Mod{\cv_{st}} \subseteq \Mod{\cv_{op}}$. This is the
model-theoretic consequence relation, written as $\cv_{st} \models_m
\cv_{op}$. In this case, we say that $\cv_{op}$ is \emph{weaker than}
$\cv_{st}$ (written $\cv_{op} \le \cv_{st}$).

Observe that our contract language (stripped of its syntactic sugar) is a
carefully chosen subset of first-order logic that is known to be
decidable~\cite{epr}.  Since first-order logic is sound and complete
~\cite{completeness}, $\cv_{st} \models_m \cv_{op}$ if and only if
$\cv_{st} \Rightarrow \cv_{op}$.  Due to the decidability of our
contract language, this implication check is automatically discharged
with the help of a theorem prover.

@@

\item \textbf{Eventually consistency}: Eventually consistent operations can
  be satisfied as long as the client can reach at least one replica. For
  example, \cf{deposit} is an eventually consistent operation; its semantics
  does not require its action to manifest on all replicas before other
  operations in its session are allowed to proceed. While eventually
  consistent data store typically offer \emph{basic} eventual consistency
  with all possible anomalies, we assume that our store provides stronger
  semantics that remain highly-available~\cite{BailisHAT,COPS}; the store
  always exposes a causal cut of the updates. This semantics can be formally
  captured in terms of the following contract definition:
\vspace{-0.6em}
\begin{smathpar}
\ecc = \forall a,b. \hbo{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff}
\end{smathpar}
\noindent where $\small \hboZ = ((\soZ \cap \sameobjZ) \cup \visZ)^+$.

\item \textbf{Causal consistency}: Operations with causally consistent
  contracts are required to see a causally consistent snapshot of the object
  state, including the actions performed on the same session.  The latter
  requirement entails that if two operations $o_1$ and $o_2$ from the same
  session are applied to two different replicas $r_1$ and $r_2$, the second
  operation cannot be discharged until the effect of $o_1$ is merged
  with $o_2$ in both $r_1$ and $r_2$. The \cf{getBalance} operation
  requires causal consistency, as it requires the operations from the
  same session to be visible, which cannot be guaranteed under
  eventual consistency. We assume that causality is only tracked
  through operations on the same object; two operations in the same
  session but on different objects are considered causally unrelated
  under this definition. Stores typically avoid tracking causality
  across objects to mitigate overheads when causality tracking is
  unnecessary. The corresponding store semantics is captured by the
  contract $\ccc$ presented in Section~\ref{sec:store_sem}.

@@

\noindent Observe that the contracts $\scc$, $\ccc$ and $\ecc$ are
themselves totally ordered with respect to the $\le$ relation: $\ecc \le
\ccc \le \scc$. This concurs with the intuition that any contract
satisfiable under $\ecc$ or $\ccc$ is satisfiable under $\scc$, and any
contract that is satisfiable under $\ecc$ is satisfiable under
$\ccc$. Nonetheless, we determine the class of a contract based on the
\emph{weakest} guarantee (among $\ecc$, $\ccc$, and $\scc$) required to
satisfy the contract. The classification scheme is presented formally 
in Figure~\ref{sem:classify}. Along with three straightforward rules
that classify contracts into consistency classes, the classification
scheme also presents a rule that judge well-formedness of a contract.
A contract is well-formed if and only if it is satisfiable under
$\scc$ - the strongest possible consistency guarantee any store can
provide. The $\le$ relation used by the rules is defined formally
below:

\begin{definition}
A contract $\cv_{op}$ is said to be weaker than $\cv_{st}$ (written $\cv_{op}
\le \cv_{st}$ ) if and only if $\Delta \vdash \cv_{st} \Rightarrow \cv_{op}$.
\begin{center}
\end{center}
\end{definition}
\vspace{-1em}
The $\Delta$ referred in the above defintion captures assumptions
about the nature of primitive relations, such as $\Rvis$ and $\Rso$
are irreflexive, and the happens-before relation $\hbZ$ is acyclic,
preventing thin-air reads. \KC{Ideally, we need to list all the axioms
somewhere.}

@@
